Posterior Retrodictive Checks

### 2024-09-25 paper/paper-papaja-revising.Rmd
additional revisions for resubmission

author: 
  - name          : "Justin Mikell"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Arizona State University - West Campus, 4701 W. Thunderbird Rd., Glendale, AZ 85306"
    email         : "jmikell@asu.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Investigation"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Derek Powell"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Formal Analysis"
      - "Methodology"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Validation"
      - "Supervision"

@@

abstract: |
  Numerous psychological findings have shown that incidental exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect.  Under predominant accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we consider whether exposure to “premise” statements affects people's truth ratings for novel “implied” statements, a pattern of findings we call the "illusory implication" effect. We argue these effects would constitute evidence for genuine belief change from incidental exposure and  identify a handful of existing findings that offer preliminary support for this claim. Building upon these, we conduct three new pre-registered experiments to further test this hypothesis, finding additional evidence that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements, including for considerably more distant implications than those previously explored. Our findings suggest that the effects of incidental exposure reach further than previously thought, with potentially consequential implications for concerns around mis- and disinformation.

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "6791"

@@

```{r}
library(tidyverse)
library(patchwork)
library(textstem)

@@

fit_imp_quiz_int <- brm(
  response ~ familiarity + implication + B_implied_true + 
    (1 + familiarity + implication|subj_id) + 
    (1 + familiarity + implication|item_pair),  # familiarity + implication + grouping (optional)
  data = df_reg %>% filter(exposure_cond=="quiz"),
  family = cumulative(),
  chains = 4,
  cores = 4,
  file="../local/fit_imp_quiz_int_max"
)

@@

@@

@@

@@

Why should incidental exposure affect later ratings of truth? Gilbert [-@gilbert1991] proposed that any idea entertained is by default first believed and must later be "unbelieved." However, predominant theoretical accounts attribute illusory truth effects not to belief but rather to metacognitive experiences of fluency or familiarity [@dechene.etal2010]. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [@alter.oppenheimer2009; @schwarz2015].

Metacognitive processing "fluency" is the experienced ease of processing or thinking about information [@alter.oppenheimer2009; @wanke.hansen2015]. When thinking is easy and fluent, people tend to judge the targets of that thinking more favorably than if thinking is challenging and disfluent [@wanke.etal1997]. A number of findings have shown that processing fluency contributes to the illusory truth effect: Incidental prior exposure to statements makes later processing of those statements more fluent, and this metacognitive experience of fluency leads people to rate those statements as more true [@difonzo.etal2016; @fazio.etal2015; @unkelbach2007; @wang.etal2016].

@@

Both theoretical accounts provide a similar overall view of the illusory truth effect: Roughly, when asked whether something is true or false, people search their memory for knowledge pertaining to belief in the statement, but they also rely on their metacognitive experiences of familiarity and fluency. Reliance on metacognitive experiences may be, at least to some extent, adaptive: Familiarity and fluency are both signs that we have “heard this somewhere before.” According to foundational theories of communication, a general assumption that communicators strive to make true utterances is a prerequisite for successful communication and social functioning [@grice1989]. Thus, if having “heard something before” suggests that someone said it, then ecologically this is a reasonable cue to truth. Psychological studies that elicit the illusory truth effect hijack these metacognitive heuristics: They provide this sense of familiarity or fluency, but from a communicative context that lacks any reasonable assumption of truth.

@@

However, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge statements expressing it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 U.S. election” might imply the belief that “Donald Trump actually won the 2020 U.S. election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

@@

Figure \@ref(fig:theory-fig) depicts our hypotheses concerning the causal relationships between latent beliefs, metacognitive experiences, and overt behaviors such truth ratings and actions. As shown in the figure, experiences or manipulations that affect metacognitive experiences, such as incidental exposure, might produce increased truth ratings for statements expressing $P$ (either through lexical or conceptual similarity). However, beliefs should be expected to produce additional consequences. Here, we focus on inferences: a manipulation affecting belief in $P$ should increase truth ratings for statements expressing some proposition $Q$ where $P$ implies $Q$, and decrease belief in $R$ where $P$ implies $\neg R$. Figure \@ref(fig:theory-fig) also depicts a possible relation between metacognitive experiences and statements expressing $Q$. Where a statement $S_Q$ expressing $Q$ invokes a similar set of concepts as $S_P$, conceptual fluency might produce increased truth ratings.

Crucially, our theory draws no connection between metacognitive experiences and reduced truth judgments for statements expressing $R$ where $P$ implies $\neg R$. Such a connection would require a cogent theory of how conceptual "disfluency" might be engendered to reduce truth judgments without the process of inference that would implicate changes in belief. Current accounts have not developed such a theory [e.g. c.f. @alter.oppenheimer2009; @VogelEtAl2020Truth], and it appears unlikely to be workable: people frequently encounter chains of statements invoking different concepts, yet they do so seemingly without issue, as considering one topic does not necessarily make it more difficult to consider another unrelated topic.

We argue that where effects of incidental exposure on truth judgments are attributable to metacognitive processes, these effects do not generally constitute evidence for a change in beliefs. Applying the theoretical framework sketched in Figure \@ref(fig:theory-fig), cases where incidental exposure to a proposition $P$ increases truth ratings for statements $S_P$ expressing that proposition do not generally constitute evidence that incidental exposure has affected beliefs, since there is a causal pathway between exposure and this result that does not pass through belief. Likewise, evidence of exposure to $S_P$ increasing truth ratings for $S_Q$ where $S_Q$ shares a similar set of concepts with $S_P$ do not constitute evidence for effects on beliefs. However, evidence that exposure decreases truth ratings for $S_R$ or increases truth ratings for $S'_Q$ where $S'_Q$ does not directly share concepts with $S_P$ would constitute evidence for effects on beliefs.

@@

A handful of existing findings provide some preliminary evidence for our hypothesized illusory implication effect. A number of researchers have found that incidental exposure affects truth ratings for novel statements contradicting previously exposed statements [@garcia-marques.etal2015; @silva.etal2017; @unkelbach.rom2017].  To our knowledge, three experiments have found evidence for this pattern of influence: Garcia-Marques (n = 58) [@garcia-marques.etal2015], Silva et al. [-@silva.etal2017] (Exp. 2, n = 45), and Unkelbach and Rom [-@unkelbach.rom2017] (Exp. 3; self-referent condition; n = 27). Such a pattern is not easily attributable to purely metacogntiive processes. Instead, several of these researchers [e.g., @garcia-marques.etal2015; @silva.etal2017] attribute these lowered truth ratings to the "meaning of the statements" rather than to conceptual fluency.

@@

Silva and colleagues [-@silva.etal2017] found that exposure to statements increased truth ratings for paraphrases of those statements that expressed the same proposition while sharing only minimal surface-level linguistic features. These researchers explained their findings as resulting from conceptual fluency produced by the initial exposure to statements: prior exposure "activates" the relevant concepts, easing with their retrieval and use in processing subsequent statements and thereby producing different metacognitive experiences. Silva and colleagues [-@silva.etal2017] also found that exposure to statements *decreased* truth ratings for rephrased contradictions of those statements. However, these findings must be qualified by the exposure context used here: Silva and colleagues asked participants to rate their interest in statements, but did not warn them that the statements could be true or false. Thus the pragmatics of the task are unclear, and participants might have believed these statements because they understood them to be presented as facts.

Unkelbach and Rom [-@unkelbach.rom2017] examined both implied and contradictory statements. They found that exposure to implied statements increased truth ratings (Exp 4.) and that statements saw a significant decrease in truth ratings following incidental exposure to their contradictions in a "self-relevant" processing condition (Exp. 4). Crucially, the results of this experiment appear to preclude any purely fluency-based account of reduced truth ratings following incidental exposure to contradictory statements: These researchers found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus even if an appropriate notion of "conceptual disfluency" could be developed, it is unlikely it could be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of incidental exposure [@garcia-marques.etal2015; @gilbert1991; @silva.etal2017; @unkelbach.rom2017].

The presently-hypothesized illusory inference effect is a key theoretical link demonstrating the connection between incidental exposure and _beliefs_, and  all of the attendant downstream consequences associated with beliefs. A handful of prior findings suggest the existence of such an effect, yet this existing evidence seems far from dispositive, being based on only a handful of small studies with sometimes inconsistent methods.

@@

Prior work tested the effect of incidental exposure on implied statements using only direct contradictions of the original statement (or paraphrases of direct contradictions). Here, we sought to further test for an "illusory implication effect" by testing the effect of incidental exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether incidental exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).

In our studies, we sought to eliminate as much as possible the role of perceptual and conceptual fluency in affecting truth judgments. These efforts are aimed at demonstrating as clearly as possible a causal pathway  between incidental exposure and effects on truth ratings through changes in belief. That is, establishing a pathway from exposure to increased truth ratings for implications $Q$ and decreased truth ratings for contradictions $R$. Compared with prior work, we developed materials involving more distant inferences that do not afford direct one-to-one mappings between shared concepts in the premise and implication statements. Thus, any reduction in truth ratings for the contradicted statements following exposure to the premise statements are not reasonably attributable to disfluency (closing the alternate pathway depicted in Figure X). For instance, one statement pair was the premise statement, “No U.S. astronauts have died since the Challenger explosion in 1986” paired with the implied statement, “The space shuttle Columbia disintegrated over Texas in 2003” (true, but contradicted by the "premise"). In contrast with prior work examining conflicting statements, there is no direct one-to-one mapping of concepts between the two statements. Instead, a chain of inferences must be developed for a conclusion to be drawn: that the space shuttle Columbia contained astronauts, that its disintegration would have killed them, and that 2003 is later than 1986.

@@

```{r fig-exp1, fig.env="figure", fig.pos = "h", fig.width=6, fig.height = 3, fig.align = "center", fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 3, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options. Bars indicate observed differences, with points and errobars indicating model predictions and 50% credible intervals."}
library(tidybayes)

make_exp_figure(df_reg %>% filter(exposure_cond=="quiz"), fit_imp_quiz_int)

@@

Somewhat surprisingly, the effects of exposure on participants’ truth ratings for the premise statements were very subtle. Our primary preregistered test compared participants’ truth ratings for the exposed premise statements at test against their ratings for another set of 8 unexposed premise statements. This analysis did not find evidence of any meaningful illusory truth effect, with a posterior estimate for the exposure parameter near zero. However, a secondary preregistered analysis comparing participants’ initial truth judgments for the premise-statements to their later truth judgments for those same statements at test did find an increase in endorsements, $\beta =$ `r report_reg_coef(fit_premise_within, "phasetest")`.

@@

@@

```{r fig-exp2, fig.env="figure", fig.pos = "t", fig.width=6, fig.height = 3, fig.align = "center", fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 2, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options. Bars indicate observed differences, with points and errobars indicating model predictions and 50% credible intervals."}

make_exp_figure(df_reg2, fit_imp_exp2)
```

@@

@@

```{r fig-exp3, fig.align='center', fig.width=6, fig.height = 3, fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 3, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options Bars indicate observed differences, with points and errorbars indicating model predictions and 50% credible intervals. Predictions were generated from a model fit excluding participants who in any way indicated suspicion about the purpose of the study."}

make_exp_figure(df_reg3, fit_imp_maximal3)

@@

@@

@@

```{r, item-fluency-print-fig, fig.align = "center", fig.cap = "Word-level overlap between premise and implication statements (after lemmatization)."}
item_fluency_fig
```

@@

We observed an “illusory implication” effect across three preregistered experiments: incidental exposure to premises (e.g. that “The Tour de France has been held every year since its inception in 1903”) influenced participant’s truth ratings for examples of their logical implications (e.g. that “The Tour de France was still held during World War I and World War II”). Unlike the illusory truth effect, this effect cannot be easily explained by the effects of familiarity or fluency on the judgment process [cf. @arkes.etal1991; @fazio.etal2015; @unkelbach2007; @wang.etal2016]: whereas these metacognitive cues would be expected to uniformly increase endorsements, participants’ truth ratings for implication statements were both increased and decreased according to the implications of the corresponding premises. Thus, it appears that incidental exposure has a genuine effect on people’s beliefs beyond simply creating a sense of fluency.

>>>>

author: 
  - name          : "Justin Mikell"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "University of California, Irvine"
    email         : "jmikell@uci.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Investigation"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Derek Powell"
    affiliation   : "2"
    address       : "Arizona State University - West Valley"
    role:
      - "Conceptualization"
      - "Formal Analysis"
      - "Methodology"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Validation"
      - "Supervision"

@@

abstract: |
  Numerous psychological findings have shown that incidental exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect.  Under many accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we consider whether exposure to “premise” statements affects people's truth ratings for novel “implied” statements, a pattern of findings we call the "illusory implication" effect. We argue these effects would constitute evidence for genuine belief change from incidental exposure and  identify a handful of existing findings that offer preliminary support for this claim. Building upon these, we conduct three new pre-registered experiments to further test this hypothesis, finding additional evidence that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements, including for considerably more distant implications than those previously explored. Our findings suggest that the effects of incidental exposure reach further than previously thought, with potentially consequential implications for concerns around mis- and disinformation.

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "7531"

@@

```{r}
library(tidyverse)
library(patchwork)
library(textstem)
library(tidybayes)

@@

fit_imp_quiz_int_dist <- brm(
  bf(response ~ familiarity + implication + B_implied_true + 
    (1 + familiarity + implication|subj_id) + 
    (1 + familiarity + implication|item_pair)) + lf(disc ~ 0 + familiarity + implication + B_implied_true),  # familiarity + implication + grouping (optional)
  data = df_reg %>% filter(exposure_cond=="quiz"),
  family = cumulative(),
  chains = 4,
  cores = 4,
  file="../local/fit_imp_quiz_int_max_dist"
)

fit_imp_quiz_int <- brm(
  bf(response ~ familiarity + implication + B_implied_true + 
    (1 + familiarity + implication|subj_id) + 
    (1 + familiarity + implication|item_pair)),
  data = df_reg %>% filter(exposure_cond=="quiz"),
  family = cumulative(),
  chains = 4,
  cores = 4,
  file="../local/fit_imp_quiz_int_max"
)

@@

fit_imp_quiz_int_bin <- brm(
  response_bin ~ familiarity + implication + B_implied_true + 
    (1 + familiarity + implication|subj_id) + 
    (1 + familiarity + implication|item_pair),  # familiarity + implication + grouping (optional)
  data = df_reg %>% filter(exposure_cond=="quiz") %>% mutate(response_bin = if_else(grepl("True", response), 1, 0)),
  family = bernoulli(),
  chains = 4,
  cores = 4,
  file="../local/fit_imp_quiz_int_max_bin"
)

@@

fit_imp_exp2_bin <- brm(
  response_bin ~ familiarity + implication + B_implied_true + 
    (1 + familiarity + implication|subj_id) + 
    (1 + familiarity + implication|item_pair),  # familiarity + implication + grouping (optional)
  data = df_reg2 %>% mutate(response_bin = if_else(grepl("True", response), 1, 0)),
  family = bernoulli(),
  prior = set_prior("lkj(3)", class = "cor"),
  chains = 4,
  cores = 4,
  iter = 3000,
  control = list(adapt_delta=.9),
  file="../local/fit_imp_exp2_max_bin"
)

@@

fit_imp_maximal3_bin <- brm(
  response_bin ~ familiarity + implication + B_implied_true + 
    (1 + familiarity + implication|subj_id) + 
    (1 + familiarity + implication|item),  # familiarity + implication + grouping (optional)
  data = df_reg3_noflag %>% mutate(response_bin = if_else(grepl("True", response), 1, 0)),
  family = bernoulli(),
  prior = set_prior("lkj(3)", class = "cor"),
  chains = 4,
  cores = 4,
  iter = 3000,
  control = list(adapt_delta=.9),
  file="../local/fit_imp_maximal3_bin"
)

@@

Why should incidental exposure affect later ratings of truth? Gilbert [-@gilbert1991] proposed that any idea entertained is by default first believed and must later be "unbelieved." This perspective offers one explanation for the illusory truth effect: people might simply believe what they are told, even when that exposure is incidental, or the pragmatics of the situation indicate that they should withhold judgment [@garcia-marques.etal2015; @silva.etal2017; @unkelbach.rom2017].

In contrast, rather than imagine people are truly so credulous, another set of theoretical accounts attribute illusory truth effects to metacognitive experiences of fluency or familiarity [@dechene.etal2010], rather than genuine influences on beliefs. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [@alter.oppenheimer2009; @schwarz2015]. Metacognitive processing "fluency" is the experienced ease of processing or thinking about information [@alter.oppenheimer2009; @wanke.hansen2015]. When thinking is easy and fluent, people tend to judge the targets of that thinking more favorably than if thinking is challenging and disfluent [@wanke.etal1997]. A number of findings have shown that processing fluency contributes to the illusory truth effect: Incidental prior exposure to statements makes later processing of those statements more fluent, and this metacognitive experience of fluency leads people to rate those statements as more true [@difonzo.etal2016; @fazio.etal2015; @unkelbach2007; @wang.etal2016].

@@

Both metacognitive accounts provide a similar overall view of the illusory truth effect: Roughly, when asked whether something is true or false, people search their memory for knowledge pertaining to belief in the statement, but they also rely on their metacognitive experiences of familiarity and fluency. Reliance on metacognitive experiences may be, at least to some extent, adaptive: Familiarity and fluency are both signs that we have “heard this somewhere before.” According to foundational theories of communication, a general assumption that communicators strive to make true utterances is a prerequisite for successful communication and social functioning [@grice1989]. Thus, if having “heard something before” suggests that someone said it, then ecologically this is a reasonable cue to truth. On this account, psychological studies that elicit the illusory truth effect hijack these metacognitive heuristics: They provide this sense of familiarity or fluency, but from a communicative context that lacks any reasonable assumption of truth.

@@

The extent of these impacts depends upon the source of the illusory truth effect, whether it is a product of metacognitive processes or results from the adoption of incidentally experienced information as beliefs. Beliefs inform actions and decision-making: When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021]. In addition, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 U.S. election” might imply the belief that “Donald Trump actually won the 2020 U.S. election.”

<!-- To be clear, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge statements expressing it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 U.S. election” might imply the belief that “Donald Trump actually won the 2020 U.S. election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].  -->

@@

Some researchers have argued that the illusory truth effect itself demonstrates an influence of prior exposure on beliefs [e.g. @garcia-marques.etal2015]. Yet as these points illustrate, there is more to _belief_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge statements expressing it to be truthful) is an important marker and convenient measure of belief, these other features are just as essential. Moreover, we argue that where effects of incidental exposure on truth judgments are plausibly attributable to well-established metacognitive processes, these effects do not generally constitute evidence for a change in beliefs.

Nevertheless, these wider consequences can help to determine whether incidental exposure to information modifies beliefs or influences judgments solely through metacognitive processes. Figure \@ref(fig:theory-fig) depicts a set of hypotheses concerning the causal relationships between latent beliefs, metacognitive experiences, and overt behaviors such truth ratings and actions. As shown in the figure, experiences or manipulations that affect metacognitive experiences, such as incidental exposure, might produce increased truth ratings for statements expressing $P$ (either through lexical or conceptual similarity). However, beliefs should be expected to produce additional consequences. Here, we focus on inferences: a manipulation affecting belief in $P$ should increase truth ratings for statements expressing some proposition $Q$ where $P$ implies $Q$, and decrease belief in $R$ where $P$ implies $\neg R$. Figure \@ref(fig:theory-fig) also depicts a possible relation between metacognitive experiences and statements expressing $Q$. Where a statement $S_Q$ expressing $Q$ invokes a similar set of concepts as $S_P$, conceptual fluency might produce increased truth ratings.

Crucially, our diagram draws no connection between metacognitive experiences and reduced truth judgments for statements expressing $R$ where $P$ implies $\neg R$. Such a connection would require a cogent theory of how conceptual "disfluency" might be engendered to reduce truth judgments without the process of inference that would implicate changes in belief. Current accounts have not developed such a theory [e.g. c.f. @alter.oppenheimer2009; @VogelEtAl2020Truth], and it appears unlikely to be workable: people frequently encounter chains of statements invoking different concepts, yet they do so seemingly without issue, as considering one topic does not necessarily make it more difficult to consider another unrelated topic.

Applying the causal model sketched in Figure \@ref(fig:theory-fig), cases where incidental exposure to a proposition $P$ increases truth ratings for statements $S_P$ expressing that proposition do not, in our view, constitute evidence that incidental exposure has affected beliefs, since there is a potential causal pathway between exposure and this result that does not pass through belief. Likewise, evidence of exposure to $S_P$ increasing truth ratings for $S_Q$ where $S_Q$ shares a similar set of concepts with $S_P$ do not constitute evidence for effects on beliefs. However, evidence that exposure decreases truth ratings for $S_R$ or increases truth ratings for $S'_Q$ where $S'_Q$ does not directly share concepts with $S_P$ would constitute evidence for effects on beliefs.

@@

A handful of existing findings provide some preliminary evidence for our hypothesized illusory implication effect. A number of researchers have found that incidental exposure affects truth ratings for novel statements contradicting previously-seen statements [@garcia-marques.etal2015; @silva.etal2017; @unkelbach.rom2017]. To our knowledge, three studies have found evidence for this specific pattern of influence: Garcia-Marques ($n = 58$) [@garcia-marques.etal2015], Silva et al. [-@silva.etal2017] (Exp. 2, $n = 45$), and Unkelbach and Rom [-@unkelbach.rom2017] (Exp. 3; $n = 89$). Such a pattern is not easily attributable to purely metacogntiive processes.
<!-- Instead, these researchers [e.g., @garcia-marques.etal2015; @silva.etal2017] attribute these lowered truth ratings to the "meaning of the statements" rather than to conceptual fluency. -->

@@

Silva and colleagues [-@silva.etal2017] found that exposure to statements increased truth ratings for paraphrases of those statements that expressed the same proposition while sharing only minimal surface-level linguistic features. These researchers explained their findings as resulting from conceptual fluency produced by the initial exposure to statements: prior exposure "activates" the relevant concepts, easing with their retrieval and use in processing subsequent statements and thereby producing different metacognitive experiences. However, Silva and colleagues [-@silva.etal2017] also found that exposure to statements *decreased* truth ratings for rephrased contradictions of those statements, which we identify as a key test of an "illusory implication" effect.

Similarly, Unkelbach and Rom [-@unkelbach.rom2017] examined both implied and contradictory statements. They found that exposure to implied statements increased truth ratings (Exp 4) and that statements saw a significant decrease in truth ratings following incidental exposure to their contradictions (Exp. 3). Crucially, the results of this experiment appear to preclude any purely fluency-based account of reduced truth ratings following incidental exposure to contradictory statements: These researchers found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus, even if an appropriate notion of "conceptual disfluency" could be developed, it is unlikely it could be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of incidental exposure [@garcia-marques.etal2015; @gilbert1991; @silva.etal2017; @unkelbach.rom2017].

However, these last two sets of findings must be qualified by the exposure context used in these studies: These researchers asked participants to make incidental judgments about these statements but did not warn them that the statements could be true or false [@silva.etal2017; @unkelbach.rom2017]. Thus, the pragmatics of the task are unclear, and participants might have believed these statements because they understood them to be presented as facts.[^urnote]

[^urnote]: Unkelbach and Rom [-@unkelbach.rom2017] did test a "warning" condition in another experiment of a similar design (Exp. 1), but here did not find a significant reduction in truth judgments for contradictions of previously-seen statements.

The presently-hypothesized illusory inference effect is a key theoretical link demonstrating the connection between incidental exposure and _beliefs_, and  all of the attendant downstream consequences associated with beliefs. A handful of prior findings suggest the existence of such an effect, yet we do not take this existing evidence to be dispositive.

@@

Existing research examining the effects of incidental exposure on implied statements has two key limitations. First, it has often been conducted in ambiguous contexts where participants might not understand that statements might be true or false, and that therefore do not always clearly induce incidental exposure. We address this through study designs inducing clear cases of incidental exposure. Second, prior work has only examined effects on direct contradictions of originally-seen statements (or paraphrases of direct contradictions). Here, we sought to further test for an "illusory implication effect" by testing the effect of incidental exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether incidental exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).

In our studies, we sought to eliminate as much as possible the role of perceptual and conceptual fluency in affecting truth judgments. These efforts are aimed at demonstrating as clearly as possible a causal pathway  between incidental exposure and effects on truth ratings through changes in belief. That is, establishing a pathway from exposure to increased truth ratings for implications $Q$ and decreased truth ratings for contradictions $R$. Compared with prior work, we developed materials involving more distant inferences that do not afford direct one-to-one mappings between shared concepts in the premise and implication statements. Thus, any reduction in truth ratings for the contradicted statements following exposure to the premise statements are not reasonably attributable to disfluency (closing the alternate pathway depicted in Figure \@ref(fig:theory-fig)). For instance, one statement pair was the premise statement, “No U.S. astronauts have died since the Challenger explosion in 1986” paired with the implied statement, “The space shuttle Columbia disintegrated over Texas in 2003” (true, but contradicted by the "premise"). In contrast with prior work examining conflicting statements, there is no direct one-to-one mapping of concepts between the two statements. Instead, a chain of inferences must be developed for a conclusion to be drawn: that the space shuttle Columbia contained astronauts, that its disintegration would have killed them, and that 2003 is later than 1986.

@@

```{r fig-exp1, fig.env="figure", fig.pos = "h", fig.width=6, fig.height = 3, fig.align = "center", fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 3, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options. Bars indicate observed differences, with points and errobars indicating binary logistic regression model predictions and 50% credible intervals."}
library(tidybayes)

make_exp_figure_two_models(df_reg %>% filter(exposure_cond=="quiz"), fit_imp_quiz_int, fit_imp_quiz_int_bin)

@@

Somewhat surprisingly, the effects of exposure on participants’ truth ratings for the premise statements were very subtle. Our primary preregistered test compared participants’ truth ratings for the exposed premise statements at test against their ratings for another set of 8 unexposed premise statements. This analysis did not find evidence of any meaningful illusory truth effect, with a posterior estimate for the exposure parameter near zero. However, an additional preregistered analysis comparing participants’ initial truth judgments for the premise-statements to their later truth judgments for those same statements at test did find an increase in endorsements, $\beta =$ `r report_reg_coef(fit_premise_within, "phasetest")`.

@@

Examination of posterior predictive plots across predictor levels indicate some modest misspecification of the model. To corroborate the primary preregistered analysis, we conducted a secondary logistic regression analysis after binarizing responses into "true" and "false" categories. The estimate of the implication effect in this model was consistent with the primary preregistered analysis $\beta =$ `r report_reg_coef(fit_imp_quiz_int_bin, "implication")`. This model's predictions are plotted in Figure \@ref(fig:fig-exp1), right.

@@

```{r fig-exp2, fig.env="figure", fig.pos = "t", fig.width=6, fig.height = 3, fig.align = "center", fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 2, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options. Bars indicate observed differences, with points and errobars indicating binary logistic regression model predictions and 50% credible intervals."}

make_exp_figure_two_models(df_reg2, fit_imp_exp2, fit_imp_exp2_bin)
```

@@

Examination of posterior predictive plots across predictor levels indicate a good fit of the ordinal regression model to participants' observed responses. Nevertheless, for consistency of presentation, we conducted a secondary logistic regression analysis after binarizing responses into "true" and "false" categories. The estimate of the implication effect in this model was consistent with the primary preregistered analysis $\beta =$ `r report_reg_coef(fit_imp_exp2_bin, "implication")`. This model's predictions are plotted in Figure \@ref(fig:fig-exp2), right.

@@

@@

```{r fig-exp3, fig.pos = "H", fig.align='center', fig.width=6, fig.height = 3, fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 3, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options. Bars indicate observed differences, with points and errobars indicating binary logistic regression model predictions and 50% credible intervals. Plotted responses and model predictions exclude participants who in any way indicated suspicion about the purpose of the study. Note that, as before, these model predictions exclude random effects of items and subject and so will not necessarily align precisely with the observed data."}

make_exp_figure_two_models(df_reg3_noflag, fit_imp_maximal3, fit_imp_maximal3_bin)

@@

As with Experiment 1, examination of posterior predictive plots across predictor levels indicate some modest misspecification of the model. To corroborate the primary preregistered analysis, we conducted a secondary logistic regression analysis after binarizing responses into "true" and "false" categories. The estimate of the implication effect in this model was generally consistent with the primary preregistered analysis $\beta =$ `r report_reg_coef(fit_imp_maximal3_bin, "implication")`. This model's predictions are plotted in Figure \@ref(fig:fig-exp3), right.

@@

```{r, item-fluency-print-fig, fig.pos = "H", fig.align = "center", fig.cap = "Word-level overlap between premise and implication statements (after lemmatization)."}
item_fluency_fig
```

@@

We observed an “illusory implication” effect across three preregistered experiments: incidental exposure to premises (e.g. that “The Tour de France has been held every year since its inception in 1903”) influenced participant’s truth ratings for examples of their logical implications (e.g. that “The Tour de France was still held during World War I and World War II”). Unlike the illusory truth effect, this effect cannot be easily explained by the effects of familiarity or fluency on the judgment process [cf. @arkes.etal1991; @fazio.etal2015; @unkelbach2007; @wang.etal2016]: whereas these metacognitive cues would be expected to uniformly increase endorsements, participants’ truth ratings for implication statements were both increased and decreased according to the implications of the corresponding premises. Thus, it appears that incidental exposure has a genuine effect on people’s beliefs beyond simply creating a sense of fluency or familiarity.
